---
title: "Statistical analysis with R for project \"Test\""
author: "Timothée Flutre"
date: '`r as.character(format(Sys.Date(), format="%d/%m/%Y"))`'
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: TRUE
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: TRUE
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(echo=TRUE, warning=TRUE, message=TRUE, cache=FALSE)#, fig.width=12, fig.height=8, fig.path=figures/")
options(digits=3)
seed <- 1859; set.seed(seed)
## suppressPackageStartupMessages(library(MASS))
```


# Preamble

For an Rmd file to be successfully exported to HTML and PDF, read [this page](http://rmarkdown.rstudio.com/).

Practically, open R and type:
```{r, eval=FALSE}
library(rmarkdown)
render("myanalysis.Rmd", "all")
```

Any Rmd file is best edited with [RStudio](http://www.rstudio.com/), or [Emacs](http://www.gnu.org/software/emacs/) with [ESS](http://ess.r-project.org/), [markdown](http://jblevins.org/projects/markdown-mode/) and [polymode](https://github.com/vitoshka/polymode).


Example of an equation written in [LaTeX](http://www.latex-project.org/): $y_i \sim \mathcal{N}(\mu, \sigma^2)$

Example of a list:

* $\mu$

* $\sigma$

Example of a plot from R:
```{r plot_hist}
hist(rnorm(10^4), breaks="FD")
```

Example of a table from R:
```{r, results='asis'}
knitr::kable(mtcars[1:5,])
```


# Overview

This document contains the statistical analysis for the project "Test".
The goal of this project is to ...


# Define the question

...


# Design the experiment

...


# Simulate some data

...


# Load and format the real data

...


# Explore the data

...


# Write the model

...


# Perform inference

...


# Check assumptions and improve the model

...


```{r info}
print(sessionInfo(), locale=FALSE)
```
