---
title: "Statistical analysis with R for project \"Test\""
author: "Timothée Flutre"
date: "01/01/2014"
output:
  html_document:
    toc: true
  pdf_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(echo=TRUE, warning=TRUE, message=TRUE, cache=FALSE)
options(digits=3)
seed <- 1859; set.seed(seed)
## suppressPackageStartupMessages(library(MASS))
```


## Preamble

For an Rmd file to be successfully exported to HTML and PDF, read [this page](http://rmarkdown.rstudio.com/).

Practically, open R and type:
```{r, eval=FALSE}
library(rmarkdown)
render("myanalysis.Rmd", "all")
```

Any Rmd file is best edited with RStudio, or Emacs with ESS, markdown and polymode.


Example of an equation written in LaTeX: $y_i \sim \mathcal{N}(\mu, \sigma^2)$

Example of a list:

* $\mu$

* $\sigma$

Example of a plot from R:
```{r plot_hist}
hist(rnorm(10^4), breaks="FD")
```

Example of a table from R:
```{r, results='asis'}
knitr::kable(mtcars[1:5,])
```


## Overview

This document contains the statistical analysis for the project "Test".
The goal of this project is to ...


## Define the question

...


## Design the experiment

...


## Simulate some data

...


## Load and format the real data

...


## Explore the data

...


## Write the model

...


## Perform inference

...


## Check assumptions and improve the model

...
