---
title: "Breeding game: setup"
author: "Timothée Flutre, Jacques David"
date: "`r format(Sys.time(), '%d/%m/%Y %H:%M:%S')`"
colorlinks: true
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: TRUE
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: TRUE
urlcolor: blue
---

<!--
This R chunk is used to set up important options and load required packages.
-->
```{r setup, include=FALSE}
R.v.maj <- as.numeric(R.version$major)
R.v.min.1 <- as.numeric(strsplit(R.version$minor, "\\.")[[1]][1])
if(R.v.maj < 2 || (R.v.maj == 2 && R.v.min.1 < 15))
  stop("requires R >= 2.15", call.=FALSE)

suppressPackageStartupMessages(library(knitr))
opts_chunk$set(echo=TRUE, warning=TRUE, message=TRUE, cache=FALSE, fig.align="center")
```


# Overview

* https://github.com/timflutre/atelier-prediction-genomique

* Copyright 2016-2017 INRA, Montpellier SupAgro

* License: AGPL-3+

This R chunk is used to assess how much time it takes to execute the R code in this document until the end:
```{r time_0}
t0 <- proc.time()
```


# Set up the environment

Load the packages:
```{r load_pkg}
library(rutilstimflutre)
library(RSQLite)
library(MASS)
## library(LDheatmap) # optionnal
```

Set up directories:
```{r setup_dir}
root.dir <- "~/work2/atelier-prog-selection-2017"
if(! dir.exists(root.dir))
  dir.create(root.dir)
## shared.dir <- "~/Dropbox/atelier-prog-selection-2016"
setup <- setUpBreedingGame(root.dir=root.dir, shared.dir=NULL,
                           nb.breeders=4, lang="fr")
truth.dir <- setup$truth.dir
shared.dir <- setup$shared.dir
init.dir <- setup$init.dir
breeders <- setup$breeders
breeder.dirs <- setup$breeder.dirs
dbname <- setup$dbname
```


# Simulate haplotypes and genotypes

Nb of initial individuals:
```{r init_inds}
I <- 1000
ind.ids <- sprintf(fmt=paste0("ind%0", floor(log10(I))+1, "i"), 1:I)
```

Simulate haplotypes and genotypes of initial individuals:

* hyp: single pop

* hyp: no inter-chrom LD

* hyp: all chroms have the same length

* hyp: individuals are diploids

* scaling: measure time in units of 4 x N_e generations

```{r init_chrs}
nb.chrs <- 10
L <- 10^6 # chromosome length, in base pairs
mu <- 10^(-8) # neutral mutation rate in events / base / generation
(u <- mu * L) # neutral mutation rate in events / chrom / gen
c <- mu # recomb rate in events / base / gen
(r <- c * L) # recomb rate in events / chrom / gen
```

Simplest evolutionary scenario similar to real data for domesticated plants:
```{r evol_scenario}
##                             _____
##                            /
##       ____________________/      Nec
##                       _Neb_
##                       |    \_____
##            Nea        |
##                       |
##                       |
##       ________________|
##                       2   1      0
## time  <---------------+---+------+
Nea <- 5 * 10^4         # ancestral
Neb <- 10 * 10^2        # bottleneck
Nec <- Ne0 <- 5 * 10^3  # current
(theta <- 4 * Ne0 * u) # scaled neutral mutation rate in events / chrom
(rho <- 4 * Ne0 * r) # scaled recomb rate in events / chrom
T1 <- 8000
T2 <- T1 + 2000
(alpha <- - (4 * Ne0) / T1 * log(Neb / Ne0)) # exp growth rate from Nec to Neb
(cmd <- paste0("-eG ", T1/(4*Ne0), " ", alpha,
               " -eN ", T2/(4*Ne0), " ", Nea/Ne0))
```

Simulate genotypes via the coalescent:
```{r simul_geno_coalescent}
set.seed(1)
f <- paste0(truth.dir, "/g0.RData")
if(! file.exists(f)){
  g0 <- simulCoalescent(nb.inds=I,
                        ind.ids=ind.ids,
                        nb.reps=nb.chrs,
                        pop.mut.rate=theta,
                        pop.recomb.rate=rho,
                        chrom.len=L,
                        other=cmd,
                        nb.pops=1,
                        get.trees=FALSE,
                        get.tmrca=FALSE,
                        verbose=1)
  save(g0, file=f)
} else
  load(f)
tools::md5sum(path.expand(f))
head(g0$snp.coords)
g0$haplos$chr1[1:4, 1:6]
g0$genos[1:2, 1:6]
hist(x=estimSnpAf(X=g0$genos), xlab="Allele frequency", ylab="Number of markers",
     main=paste0(ncol(g0$genos), " SNPs"), col="grey", border="white", las=1, breaks="FD")
```

Look at the MAFs:
```{r look_mafs}
mafs <- estimSnpMaf(X=g0$genos)
summary(mafs)
sum(mafs >= 0.01) # nb of non-rare SNPs
plotHistMinAllelFreq(maf=mafs,
                     main=paste0(length(mafs), " SNPs"))
plotHistMinAllelFreq(maf=mafs[mafs >= 0.01],
                     main=paste0(sum(mafs >= 0.01), " SNPs (MAF >= 0.01)"))
```

Look at some haplotypes:
```{r look_haplos}
plotHaplosMatrix(g0$haplos$chr1[sort(sample.int(nrow(g0$haplos$chr1), 100)),
                                mafs[1:ncol(g0$haplos$chr1)] >= 0.2])
## calcAvgPwDiffBtwHaplos(g0$haplos$chr1) # slow
```

Estimate and plot pairwise LD per chromosome:
```{r estim_plot_ld}
chr <- "chr1"
min.maf <- 0.2
min.pos <- 0
max.pos <- L
(length(snps.tokeep <- rownames(g0$snp.coords[mafs >= min.maf &
                                               g0$snp.coords$chr == chr &
                                               g0$snp.coords$pos >= min.pos &
                                               g0$snp.coords$pos <= max.pos,])))
## snps.tokeep <- snps.tokeep[sort(sample(snps.tokeep, 100))]
(nrow(ld <- estimLd(X=g0$genos[,snps.tokeep],
                    snp.coords=g0$snp.coords[snps.tokeep,],
                    use.ldcorsv=FALSE)))
summary(ld$cor2)
snp.dist <- distSnpPairs(snp.pairs=ld[, c("loc1","loc2")],
                         snp.coords=g0$snp.coords[snps.tokeep,])
plotLd(snp.dist,
       ## ld$cor2, estim="r2",
       sqrt(ld$cor2), estim="r",
       main=paste0(length(snps.tokeep), " SNPs with MAF >= ", min.maf,
                   " on ", chr),
       use.density=TRUE,
       span=1/20,
       sample.size=2*I,
       Ne=Ne0, c=c)
abline(v=500, lty=2)
## LDheatmap(gdat=cor(g0$genos[,snps.tokeep])^2,
##           genetic.distances=snp.dist, distances="physical",
##           title=paste0(length(snps.tokeep), " SNPs with MAF >= ", min.maf,
##                        " on ", chr),
##           add.map=FALSE, flip=FALSE)
```

Look at distances between consecutive SNPs:
```{r dist_btw_snps}
chr <- "chr3"
tmp <- distConsecutiveSnps(snp.coords=g0$snp.coords, only.chr=chr)
hist(tmp[[chr]], breaks="FD", xlab="in bp", las=1,
     main="Distances between consecutive SNPs")
abline(v=500, lty=2)
```

Thin SNPs:
```{r thin}
thin.snps <- thinSnps(method="coord", threshold=5*10^4,
                      snp.coords=g0$snp.coords[names(mafs)[mafs >= 0.1],])
length(thin.snps)
```

Estimate additive genomic relationships:
```{r estim_genrel}
A.vr <- estimGenRel(X=g0$genos[,thin.snps], method="vanraden1")
A <- A.vr
imageWithScale(z=A, main="Additive genomic relationships (A)")
inbred.coefs <- diag(A) - 1
summary(inbred.coefs) # compare to 0 if no inbreeding
hist(inbred.coefs, breaks="FD")
abline(v=0, lty=2, col="red")
coancestries <- A / 2
summary(diag(coancestries)) # compare to 0.5 for an outbred individual
hist(diag(coancestries), breaks="FD", xlim=c(0,1))
abline(v=0.5, lty=2, col="red")
summary(coancestries[upper.tri(coancestries)])
hist(coancestries[upper.tri(coancestries)], breaks="Sturges")
abline(v=0, lty=2, col="red")
```

Ascertain SNPs for the genotyping arrays (high and low density):

* hyp: mimick de novo sequencing of a few individuals

```{r ascertain_snps}
set.seed(8)
if(! file.exists(paste0(init.dir, "/snp_coords_hd.txt.gz"))){
  nb.inds.denovo <- 20
  nb.snps.hd <- 10*10^3
  nb.snps.ld <- 5*10^3
  subset.inds <- sort(sample(rownames(g0$genos), nb.inds.denovo))
  mafs.subset <- estimSnpMaf(g0$genos[subset.inds,])
  summary(mafs.subset)
  sum(mafs.subset >= 0.1)
  subset.snps <- list()
  subset.snps[["hd"]] <- sample(names(mafs.subset[mafs.subset >= 0.1]), nb.snps.hd)
  subset.snps[["ld"]] <- sample(subset.snps[["hd"]], nb.snps.ld)
  summary(mafs[subset.snps[["hd"]]])
  summary(mafs[subset.snps[["ld"]]])
  f <- paste0(init.dir, "/snp_coords_hd.txt.gz")
  write.table(x=g0$snp.coords[subset.snps[["hd"]],],
              file=gzfile(f), quote=FALSE,
              sep="\t", row.names=TRUE, col.names=TRUE)
  f <- paste0(init.dir, "/snp_coords_ld.txt.gz")
  write.table(x=g0$snp.coords[subset.snps[["ld"]],],
              file=gzfile(f), quote=FALSE,
              sep="\t", row.names=TRUE, col.names=TRUE)
} else{
  subset.snps <- list()
  f <- paste0(init.dir, "/snp_coords_hd.txt.gz")
  subset.snps[["hd"]] <- rownames(read.table(f))
  f <- paste0(init.dir, "/snp_coords_ld.txt.gz")
  subset.snps[["ld"]] <- rownames(read.table(f))
}
```

Make collection of lines via haplodiploidization:
```{r coll_lines}
set.seed(5)
f <- paste0(truth.dir, "/coll.Rdata")
if(! file.exists(f)){
  coll <- list()
  crosses <- data.frame(parent1=ind.ids,
                        parent2=NA,
                        child=sprintf(fmt=paste0("Coll%0", floor(log10(I))+1, "i"),
                                      1:I),
                        stringsAsFactors=FALSE)
  loc.crossovers <- drawLocCrossovers(crosses, sapply(g0$haplos, ncol))
  coll$haplos <- makeCrosses(g0$haplos, crosses, loc.crossovers)
  coll$genos <- segSites2allDoses(seg.sites=coll$haplos, ind.ids=crosses$child,
                                  snp.ids=colnames(g0$genos),
                                  rnd.choice.ref.all=FALSE)
  coll$snp.coords <- g0$snp.coords
  save(coll, file=f)
} else
  load(f)
tools::md5sum(path.expand(f))
```

Free some memory:
```{r free_mem_genos}
## sort(sapply(ls(),function(x){object.size(get(x))}), decreasing=TRUE)
rm(g0, ld, snp.dist, mafs, loc.crossovers, mafs.subset, crosses,
   coancestries, A, A.vr)
gc()
```


# Simulate phenotypes

## Generic preparation

Make data.frame for initial phenotypes given to players:
```{r init_pheno}
It <- 150      # nb of lines per trial (i.e. year and site)
J <- 10        # years
K <- 1         # site
L <- 300       # plots per site
R <- 2         # replicate per line per trial
N <- J * K * L # nb of measurements per trait for all trials
first.year <- 2005
latest.year <- first.year + J - 1
p0.df <- data.frame(ind=NA,
                    year=as.factor(rep(first.year:latest.year, each=L)),
                    plot=as.factor(rep(1:L, times=J)),
                    pathogen=NA,
                    trait1.raw=NA,
                    trait1=NA,
                    trait2=NA,
                    trait3=NA)
It.new <- 75   # nb of new lines per year (starting the second year)
line.ids <- getIndNamesFromHaplos(coll$haplos)[1:(It+ (J-1) * It.new)]
(length(line.ids)) # total nb of lines tested at least one year
p0.df$ind[p0.df$year == levels(p0.df$year)[1]] <- rep(line.ids[1:It],
                                                      each=R)
for(j in 1:J){
  year <- levels(p0.df$year)[j]
  p0.df$ind[p0.df$year == year] <- rep(line.ids[(1+(j-1)*It.new):((j-1)*It.new+It)],
                                       each=R)
}
p0.df$ind <- as.factor(p0.df$ind)
head(p0.df)
str(p0.df)
```

hyp: pathogens present (and detected) every three years:
```{r patho_years}
p0.df$pathogen <- FALSE
years.with.patho <- seq(from=first.year, by=3, length.out=4)
p0.df$pathogen[p0.df$year %in% years.with.patho] <- TRUE
```

Export genotypes for the players using high-density chip, but only for lines planted the last year:
```{r export_geno}
f <- paste0(init.dir, "/genos_coll-", latest.year, ".txt.gz")
subset.coll <- unique(as.character(p0.df$ind[p0.df$year == latest.year]))
write.table(x=coll$genos[subset.coll, subset.snps[["hd"]]],
            file=gzfile(f), quote=FALSE,
            sep="\t", row.names=TRUE, col.names=TRUE)
```

Simulate phenotypes of initial lines:

* hyp: only 3 traits are considered

* hyp: trait 1 has additive infinitesimal genetic architecture (e.g. yield)

* hyp: trait 2 has additive infinitesimal genetic architecture (e.g. protein)

* hyp: traits 1 and 2 are negatively genetically correlated (i.e. pleiotropy)

* hyp: for traits 1 and 2, all SNPs (even the rarest) have a non-zero effect

* hyp: trait 3 has a single, major QTL (e.g. resistance)

* hyp: "year" and "plot" effects are the same for traits 1 and 2

* for traits 1 and 2, use the "equivalent" model AM <=> RR

```{r init_list_phenos}
p0 <- list(N=nrow(p0.df),
           I=nlevels(p0.df$ind),
           J=nlevels(p0.df$year),
           K=nlevels(p0.df$plot),
           Z.I=model.matrix(~ p0.df$ind - 1),
           Z.J=model.matrix(~ p0.df$year - 1),
           Z.K=model.matrix(~ p0.df$plot - 1),
           sigma.alpha2=10, # large "year" effect
           sigma.gamma2=1) # small "plot" effect
```

Thin to avoid LD between causal SNPs?
```{r}
## mafs <- estimSnpMaf(X=coll$genos[levels(p0.df$ind),])
## thin.snps <- thinSnps(method="coord", threshold=5*10^2,
##                       snp.coords=coll$snp.coords[names(mafs)[mafs >= 0.01],])
thin.snps <- colnames(coll$genos)
(p0$P <- length(thin.snps))
p0$X <- coll$genos[levels(p0.df$ind), thin.snps]
p0$afs <- estimSnpAf(X=p0$X)
```

Draw "year" and "plot" effects:
```{r draw_fix_eff}
set.seed(736452)
(p0$alpha <- setNames(rnorm(n=p0$J, mean=0, sd=sqrt(p0$sigma.alpha2)),
                      levels(p0.df$year)))
p0$gamma <- setNames(rnorm(n=p0$K, mean=0, sd=sqrt(p0$sigma.gamma2)),
                     levels(p0.df$plot))
summary(p0$gamma)
```

## Traits 1 and 2

Simulate the random part of traits 1 and 2 jointly (caution, keep `center.genos=FALSE`):
```{r make_trait1}
set.seed(39452)
traits12 <- simulTraits12Rnd(h2=c(0.3, 0.4), sigma.beta2=c(10^(-5), 10^(-5)),
                             prop.pleio=0.4, cor.pleio=-0.7,
                             X=p0$X, center.genos=FALSE, plot=TRUE)
```

Make traits 1 and 2:
```{r}
p0$trait1 <- list(beta=traits12$Beta[,1],
                  sigma2=traits12$sigma2[1],
                  u=traits12$G.A[,1])
p0$trait1$y <- p0$Z.J %*% p0$alpha + p0$Z.K %*% p0$gamma +
  p0$Z.I %*% traits12$Y[,1]
(p0$trait1$mu <- 40 - mean(p0$trait1$y))
p0$trait1$y <- p0$trait1$y + p0$trait1$mu
p0$trait2 <- list(beta=traits12$Beta[,2],
                  sigma2=traits12$sigma2[2],
                  u=traits12$G.A[,2])
p0$trait2$y <- p0$Z.J %*% p0$alpha + p0$Z.K %*% p0$gamma +
  p0$Z.I %*% traits12$Y[,2]
(p0$trait2$mu <- 14 - mean(p0$trait2$y))
p0$trait2$y <- p0$trait2$y + p0$trait2$mu
hist(p0$trait1$y, breaks="FD", las=1)
boxplot(p0$trait1$y ~ p0.df$year, las=1, main="p0$trait1$y"); abline(h=mean(p0$trait1$y), lty=2)
hist(p0$trait2$y, breaks="FD", las=1)
boxplot(p0$trait2$y ~ p0.df$year, las=1, main="p0$trait2$y"); abline(h=mean(p0$trait2$y), lty=2)
regplot(p0$trait1$y, p0$trait2$y, xlab="p0$trait1$y", ylab="p0$trait2$y")
```

Free some memory:
```{r}
rm(traits12)
gc()
```

## Trait 3

Make trait 3:

* hyp: "y=1" means "presence of symptoms" and, very likely, non-resistance

* hyp: the causal SNP is on the low-density chip

* hyp: the resistance-confering allele is the minor one, but isn't too rare

* hyp: the resistance-confering allele has a dominant effect

```{r make_trait3}
set.seed(83544)
p0$trait3 <- list(mu=0, h2=0.975)
mafs <- estimSnpMaf(afs=p0$afs)
(p0$trait3$qtn.id <- names(sample(x=which(mafs >= 0.14 & mafs <= 0.16 &
                                          names(mafs) %in% subset.snps[["ld"]]),
                                  size=1)))
mafs[p0$trait3$qtn.id]
table(p0$X[, p0$trait3$qtn.id])
(is.minor.counted <- p0$afs[p0$trait3$qtn.id] < 0.5)
if(is.minor.counted){
  p0$trait3$inds.resist <- rownames(p0$X)[p0$X[, p0$trait3$qtn.id] >= 1]
  p0$trait3$resist.genos <- c(1,2)
} else{
  p0$trait3$inds.resist <- rownames(p0$X)[p0$X[, p0$trait3$qtn.id] <= 1]
  p0$trait3$resist.genos <- c(0,1)
}
length(p0$trait3$inds.resist)
p0$trait3$y <- rep(0, p0$N) # "0" means "no symptoms"
p0$trait3$y[p0.df$pathogen] <- 1
idx <- which(p0.df$pathogen & (p0.df$ind %in% p0$trait3$inds.resist))
p0$trait3$y[idx] <- 1 - rbinom(n=length(idx), size=1, prob=p0$trait3$h2)
table(p0$trait3$y[idx])
all(unique(p0.df$ind[idx]) %in% p0$trait3$inds.resist)
length(unique(p0.df$ind[idx]))
```

## Finalize

Fill "phenotype" data.frame for players:
```{r fill_pheno_players}
p0.df$trait1.raw <- p0$trait1$y[,1]
p0.df$trait2 <- p0$trait2$y[,1]
p0.df$trait3 <- p0$trait3$y
```

Adjust trait 1 for yield loss in case of pathogen presence:
```{r adjust_trai1}
p0$prop.yield.loss <- 0.2
p0.df$trait1 <- p0.df$trait1.raw
idx <- p0.df$pathogen & (p0.df$trait3 == 1)
p0.df$trait1[idx] <- (1 - p0$prop.yield.loss) * p0.df$trait1[idx]
hist(p0.df$trait1, breaks="FD", las=1)
boxplot(p0.df$trait1 ~ p0.df$year, las=1, main="p0.df$trait1")
abline(h=mean(p0.df$trait1), lty=2)
```

Save data:
```{r save_phenos}
f <- paste0(truth.dir, "/p0.RData")
save(p0, file=f)
tools::md5sum(path.expand(f))
```

Export "phenotype" data.frame for players:
```{r export_phenos}
f <- paste0(init.dir, "/phenos_coll.txt.gz")
write.table(x=p0.df[, -grep("raw", colnames(p0.df))],
            file=gzfile(f), quote=FALSE,
            sep="\t", row.names=FALSE, col.names=TRUE)
tools::md5sum(path.expand(f))
```


# Choose controls

At the end of the game, each breeder has to submit at least one new variety.
For them to be registered, they should be "better" than controls.
The controls are chosen so that they have (i) a "genotypic yield" among the 80% best, (ii) a "genotypic quality" above 14/1000, and (iii) the resistant allele.

Rank the genotypes:
```{r}
genovals <- data.frame(trait1=p0$trait1$mu + (p0$X - 1) %*% p0$trait1$beta,
                       trait2=p0$trait2$mu + (p0$X - 1) %*% p0$trait2$beta,
                       trait3=as.numeric(! p0$X[, p0$trait3$qtn.id] %in%
                                         p0$trait3$resist.genos),
                       row.names=rownames(p0$X))
mean(genovals$trait1)
var(genovals$trait1)
mean(genovals$trait2)
var(genovals$trait2)
table(genovals$trait3)
genovals <- genovals[order(genovals$trait1, decreasing=TRUE),]
```

Identify those that can be controls:
```{r}
nrow(potential.controls <- genovals[genovals$trait2 >= 14 &
                                    genovals$trait3 == 0,])
nrow(potential.controls <- potential.controls[1:floor((1-0.8)*nrow(potential.controls)),])
```

Select five controls among them (the least good, to be easier for the players):
```{r}
nb.controls <- 5
(controls <- rownames(potential.controls)[
  (nrow(potential.controls)-nb.controls+1):nrow(potential.controls)])
genovals[controls,]
mean(genovals[controls, "trait1"])
var(genovals[controls, "trait1"])
mean(genovals[controls, "trait2"])
var(genovals[controls, "trait2"])
print(p0$X[controls, p0$trait3$qtn.id])
```

Save controls for the players:
```{r}
f <- paste0(init.dir, "/temoins.txt")
write.table(x=controls, file=f, quote=FALSE, sep="\t", row.names=FALSE, col.names=FALSE)
```

<!--
TODO: Make and save phenotypes of controls in several plots the last few years to give a kind of reference to the players:
```{r phenos_controls}
I.ctl <- length(controls) # nb of controls
J.ctl <- 4 # nb of years, from the latest
L.ctl <- 4 # nb of plots per site
first.year.ctl <- latest.year - J.ctl + 1
latest.year.ctl <- first.year.ctl + J.ctl - 1
first.plot.ctl <- L + 1
latest.plot.ctl <- first.plot.ctl + L.ctl - 1
phectl.df <- data.frame(
    ind=rep(rep(controls, each=L.ctl), J.ctl),
    year=as.factor(rep(first.year.ctl:latest.year.ctl, each=I.ctl * L.ctl)),
    plot=as.factor(rep(rep(first.plot.ctl:latest.plot.ctl, I.ctl),
                       J.ctl)),
    pathogen=FALSE,
    trait1.raw=NA,
    trait1=NA,
    trait2=NA,
    trait3=NA)
phectl.df$pathogen[phectl.df$year %in% years.with.patho] <- TRUE
```
-->

Free some memory:
```{r free_mem_phenos}
rm(p0)
gc()
```


# Set up database and example files

Create a database with tables for logging and crosses per breeder:
```{r setup_db}
db <- dbConnect(SQLite(), dbname=dbname)
dbListTables(db)
tbl <- "log"
if(tbl %in% dbListTables(db)){
  query <- paste0("DROP TABLE ", tbl)
  res <- dbGetQuery(conn=db, query)
}
query <- paste0("CREATE TABLE ", tbl,
                "(breeder TEXT",
                ", task TEXT",
                ", quantity INT",
                ", time TIMESTAMP DEFAULT CURRENT_TIMESTAMP NOT NULL)")
res <- dbGetQuery(conn=db, query)
dbListFields(db, tbl)
for(breeder in breeders){
  tbl <- paste0("crosses_", breeder)
  if(tbl %in% dbListTables(db)){
    query <- paste0("DROP TABLE ", tbl)
    res <- dbGetQuery(conn=db, query)
  }
  query <- paste0("CREATE TABLE ", tbl,
                  " (parent1 TEXT, parent2 TEXT, child TEXT PRIMARY KEY)")
  res <- dbGetQuery(conn=db, query)
}
dbListTables(db)
```

For each individual initially given to the players, save it into the table, and save its haplotypes into a file (slow, especially the first time):
```{r fill_table}
f <- paste0(truth.dir, "/coll.Rdata")
load(f)
ind.ids <- rownames(coll$genos)
tmp.ids <- gsub("Coll", "Ind", ind.ids) # dummy identifiers for the parent1's
for(i in 1:length(ind.ids)){
  ind.id <- ind.ids[i]
  ind <- list(haplos=getHaplosInd(haplos=coll$haplos, ind.name=ind.id))
  f <- paste0(truth.dir, "/", ind.id, "_haplos.RData")
  save(ind, file=f)
  for(breeder in breeders){
    fto <- paste0(truth.dir, "/", breeder, "/", ind.id, "_haplos.RData")
    if(! file.exists(fto))
      file.symlink(f, fto)
    tbl <- paste0("crosses_", breeder)
    ## print(c(breeder, ind.id))
    query <- paste0("INSERT INTO ", tbl, " VALUES",
                    "('", tmp.ids[i],
                    "', '", NA,
                    "', '", ind.id, "')")
    res <- dbGetQuery(conn=db, query)
  }
}
```

Make an example of a "cross" file:
```{r ex_cross}
(cross.examples <- makeExamplePlantFile(init.dir))
```

Make an example of an "inds" file to geno and/or pheno:
```{r ex_data}
(data.examples <- makeExampleDataFile(init.dir))
```


# Appendix

```{r info}
t1 <- proc.time(); t1 - t0
print(sessionInfo(), locale=FALSE)
```
